---
layout: page
title: Programming with R
subtitle: Addressing, merging, and sorting data
minutes: 20
---


```{r, include = FALSE}
source('tools/chunk-options.R')
```

> ## Learning Objectives {.objectives}
>
> * Understand the 3 different ways R can address data inside a data frame.
> * Combine different methods for addressing data with the assignment operator to update subsets of data
> * Merge two data frames using a shared column
> * Sort a data frame

R is a powerful language for data manipulation. There are 3 main ways for addressing data inside R objects.

* By index (slicing)
* By logical vector
* By name (columns only)

Lets start by loading some sample data:

```{r readData}
info <- read.csv(file="data/motor-info.csv")
```

Lets take a look at this data.

```{r classDat}
class(info)
```

R has loaded the contents of the .csv file into a variable called `info` which is a `data frame`.

```{r dimDat}
dim(info)
nrow(info)
ncol(info)
```

The data has 32 rows and 11 columns.

```{r headDat}
head(info)
```

The `str` function gives a nice summary:

```{r}
str(info)
```

The data are features of different makes of cars. (This is actually a standard R example data set called mtcars.)

### Addressing by Index

Data can be accessed by index. We have already seen how square brackets `[` can be used to subset (slice) data. The generic format is `info[row_numbers,column_numbers]`.

> ## Challenge - Selecting values 1 {.challenge}
>
> What will be returned by `info[1,1]`?

```{r indexing1}
info[1,1]
```

If we leave out a dimension R will interpret this as a request for all values in that dimension.

> ## Challenge - Selecting values 2 {.challenge}
>
> What will be returned by `info[,2]`?

The colon `:` can be used to create a sequence of integers.

```{r colonOperator}
6:9
```

Creates a vector of numbers from 6 to 9.

This can be very useful for addressing data.

> ## Challenge - Subsetting with sequences {.challenge}
> Use the colon operator to index just the aneurism count data (columns 6 to 9).

Finally we can use the `c()` (combine) function to address non-sequential rows and columns.

```{r nonsequential_indexing}
info[c(1,5,7,9),1:5]
```

Returns the first 5 columns for cars in rows 1,5,7 & 9

> ## Challenge - Subsetting non-sequential data {.challenge}
> Return the Age and Gender values for the first 5 patients.

### Addressing by column name

Columns in an R data frame are named.

```{r column_names}
colnames(info)
```

> ## Tip {.callout}
>
> If names are not specified e.g. using `headers=FALSE` in a `read.csv()` function, R assigns default names `V1,V2,...,Vn`

We usually use the `$` operator to address a column by name

```{r named_addressing}
info$cyl
```

Named addressing can also be used in square brackets.
```{r names_addressing_2}
head( info[,c('cyl','hp')] )
```

> ## Best Practice {.callout}
>
> Best practice is to address columns by name, often you will create or delete columns and the column position will change.

### Logical indexing

A logical vector contains only the special values `TRUE` & `FALSE`.

```{r logical_vectors}
c(TRUE,TRUE,FALSE,FALSE,TRUE)
```
> ## Tip {.callout}
>
> Note the values `TRUE` and `FALSE` are all capital letters and are not quoted.

Logical vectors can be created using `relational operators` e.g. `<, >, ==, !=, %in%`.

```{r logical_vectors_example}
x <- c(1,2,3,11,12,13)
x < 10

x %in% 1:10
```

We can use logical vectors to select data from a data frame.

```{r logical_vectors_indexing}
index <- info$cyl == 6
info[index,]$hp
```

Often this operation is written as one line of code:

```{r logical_vectors_indexing2}
plot(info[info$cyl==6,]$hp)
```


### Combining Indexing and Assignment

The assignment operator `<-` can be combined with indexing.

```{r indexing and assignment}
x<-c(1,2,3,11,12,13)
x[x < 10] <- 0
x
```


## Merging data

The magazine Motor Trends measured the miles-per-gallon of each of these cars. We want to merge this with our existing data frame.

```{r}
mpg <- read.csv("data/motor-mpg.csv")
head(mpg)
dim(mpg)
```

Hmm. This data frame has slightly less rows, and they're in a different order.

```{r}
mpg$make %in% info$make
all(mpg$make %in% info$make)
```

The `merge` function can be used to merge the two data frames:

```{r}
combined <- merge(info, mpg, by="make")
head(combined)
dim(combined)
```

We see we have only 30 rows. If we want to not discard rows that can't be merged, we can specify `all=TRUE` (or possibly `all.x=TRUE` or `all.y=TRUE`).

```{r}
combined_all <- merge(info, mpg, by="make", all=TRUE)
dim(combined_all)
combined_all
```

`NA` is a special value R uses to indicate missing data.

We might wish to save the combined data frame to a new CSV file.

```{r, eval=FALSE}
write.csv(combined_all, "combined.csv", row.names=FALSE)
```

We can use the plotting functions we learned about in M1 to examine this data.

```{r}
library(ggplot2)
ggplot(combined_all, aes(x=hp,y=mpg,color=cyl)) + geom_point() + theme_bw()
```

## Sorting data

Suppose we wanted to order rows, for example by number of cylinders. The `sort` function sorts a vectors, so we could:

```{r}
sort(combined_all$cyl)
```

However we don't just want a sorted version of this column, we want all the columns. The `order` function gives the order required to sort a vector:

```{r}
order(combined_all$cyl)
```

So we can use this to sort the data frame:

```{r}
combined_all[ order(combined_all$cyl), ]
```

> ## Challenge {.challenge}
>
> Show the top 5 best cars in terms of miles-per-gallon.
> (Look at the documentation for order to see how to sort in descending order, `?order`)
>
> What does order do with missing values (NA)?
>
